Bayesian Parameter Inference for 1D Nonlinear Stochastic Differential Equation Models

### 2015-09-16 TeXFiles/Manuscript_Albert_Ulzega.tex
Version submitted to NJP

\begin{abstract}
Bayesian statistics has become an indispensable tool in many applied sciences, in particular, for the purpose of uncertainty analysis.
Inferring parametric uncertainty, for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
%Here we present a Monte Carlo-based method for Bayesian parameter inference with generic one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
Here, we consider the generic problem of calibrating a one dimensional SDE model to time series and quantifying the ensuing parametric uncertainty.
We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to sample from it.
Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems. Furthermore the algorithm is highly parallelizable.

@@

Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data. Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}. For the case of linear SDEs that are coupled to nonlinear deterministic ordinary differential equations (ODEs), more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
The problem with these simulation-based methods is their inefficiency in the presence of many data points.
One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
Such {\em Approximate Bayes Computations} \cite{marin_2012_ABC, Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator. However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.

@@

Tuning of HMC algorithms is a non-trivial matter.
Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the {\em numerical integration} scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
Here we explore a different, computationally simper route.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
At least for 1D SDEs we always find a parametrization, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.

@@

Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
Furthermore, by means of the transformation $S(t)=1/n(t)$, eq. (\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{fujisaka_1986_intermittency}.

@@

Before we set out to derive the likelihood function, from the model equations (\ref{sde}) through (\ref{data}), let us express the parameters and state variable with dimensionless quantities.
The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}
For our algorithm it is important to have the model equation in a form where the noise term does neither depend on the state variables nor the parameters that have to be inferred. In a one dimensional model this can always be achieved through re-parametrization.

@@

Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{dutre_1977_SDE, fujisaka_1986_intermittency} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model. However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$. After plugging (\ref{pathint}) and (\ref{action}) into the detailed balance condition,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,,
\end{equation}
and using transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ onto paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}\label{equilibrium_mean}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,
\end{equation}
and diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as the path-integral
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
The first term in the exponent describes the log-probability distribution of model outputs, given model parameters, inputs and a system realization $q(t_s)$. The second term is the log-probability of a system realization $q(t)$ and the path-integral extends over all possible system realizations. Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.

@@

In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) can then be expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p(t)$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate. Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense. Thus, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.

@@

We use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
\end{enumerate}
Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is not too large and/or the measurement error $\sigma$ is not too small.
The generalization of the method to the other schemes is straightforward.
In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator for the associated Hamilton equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes much faster dynamics than the inner one. However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m'/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by the equations,
\begin{eqnarray}
  u_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{eqnarray}
for $s=1,\dots,n$ and $k=2,\dots,j$.

For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{swope_1982_verlet}.
For the boundary beads, it reads
\begin{eqnarray}
  u_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2M}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{eqnarray}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
For the model parameters $\bt$ and their momenta ${\pmb\pi}$, analogous equations apply.
When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
Thus,
\begin{eqnarray}
  p_{(s-1)j+k}(\Delta\tau) \nonumber \\
  =
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,,
\end{eqnarray}
with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over the time $\tau$.
After $P$ steps, the proposed configuration, $(\vc u',\bt';\vc p',{\pmb\pi}')$ is accepted with the Metropolis probability
\begin{equation}
  \min
  \left(
    1,
    e^{\mathcal{H}_{\mbox{\tiny HMC}}(\vc u,\bt;\vc p,{\pmb\pi}) - \mathcal{H}_{\mbox{\tiny HMC}}(\vc u',\bt';\vc p',{\pmb\pi}')}
  \right)\,,
\end{equation}
and the next iteration starts with sampling an new momentum vector $(\vc p,{\pmb\pi})$.

@@

>>>>

\begin{abstract}
Bayesian statistics has become an indispensable tool in many applied sciences, in particular, for the purpose of uncertainty analysis.
Inferring parametric uncertainty, for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
Here, we consider the generic problem of calibrating a one dimensional SDE model to time series and quantifying the ensuing parametric uncertainty.
We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to sample from it.
Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems. Furthermore the algorithm is highly parallelizable.

@@

Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density, for the parameters, is a {\em path-integral}.
Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data. Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}. For the case of linear SDEs that are coupled to nonlinear deterministic ordinary differential equations (ODEs), more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
The problem with these simulation-based methods is their inefficiency in the presence of many data points.
One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
Such {\em Approximate Bayes Computations} \cite{marin_2012_ABC, Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator. However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.

@@

Tuning of HMC algorithms is a non-trivial matter.
Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the {\em numerical integration} scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
Here we explore a different, computationally simper route.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
Thus, we employ a multiple time scale integration technique, for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
At least for 1D SDEs we always find a parametrization, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.

@@

Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
Furthermore, by means of the transformation $S(t)=1/n(t)$, eq. (\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.

@@

Before we set out to derive the likelihood function, from the model equations (\ref{sde}) through (\ref{data}), let us express the parameters and state variable with dimensionless quantities.
The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}
A one dimensional model always allows for such a parametrization, in which the noise term does neither depend on the state variables nor the parameters that have to be inferred.
This will allow us to do part of the integration later on analytically.

@@

Properties of (a transformed version of) (\ref{sde}), for constant input, have been derived in \cite{dutre_1977_SDE, schenzle_1979_multStochProc, fujisaka_1986_intermittency} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model. However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$. After plugging (\ref{pathint}) and (\ref{action}) into the detailed balance condition,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,,
\end{equation}
and using transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ onto paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}\label{equilibrium_mean}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,
\end{equation}
and diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as the path-integral
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
The first term in the exponent describes the log-probability distribution of model outputs, given model parameters, inputs and a system realization $q(t_s)$. The second term is the log-probability of a system realization $q(t)$ and the path-integral extends over all possible system realizations. In the next section, we apply a Hamiltonian Monte Carlo method to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.

@@

In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is rewritten as the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p(t)$ and ${\pmb\pi}$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate. Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense. Thus, after cutting off a possible burn-in, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.

@@

We use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
\begin{enumerate}
  \item[\it i.]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
  \item[\it ii.]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
  \item[\it iii.]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
\end{enumerate}
Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is not too large and/or the measurement error $\sigma$ is not too small.
The generalization of the method to the other schemes is straightforward.
In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
In order to design a reversible integrator, for the associated Hamilton equations, we define the Liouville operators
\begin{equation}
  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
  iL'=\{\cdot\,,\,\mathcal H'\}\,,
\end{equation}
where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes much faster dynamics than the inner one. However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m'/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The fast outer propagator is then given by the equations,
\begin{eqnarray}
  u_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
  p_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
  = p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{eqnarray}
for $s=1,\dots,n$ and $k=2,\dots,j$.

For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{swope_1982_verlet}.
For the boundary beads, it reads
\begin{eqnarray}
  u_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2M}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau) \nonumber \\
  = p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{eqnarray}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
For the model parameters $\bt$ and their momenta ${\pmb\pi}$, analogous equations apply.
When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
Thus,
\begin{eqnarray}
  p_{(s-1)j+k}(\Delta\tau) \nonumber \\
  =
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,,
\end{eqnarray}
with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over time $\tau$.
After $P$ steps, the proposed configuration, $(\vc u',\bt';\vc p',{\pmb\pi}')$ is accepted with the Metropolis probability
\begin{equation}
  \min
  \left(
    1,
    e^{\mathcal{H}_{\mbox{\tiny HMC}}(\vc u,\bt;\vc p,{\pmb\pi}) - \mathcal{H}_{\mbox{\tiny HMC}}(\vc u',\bt';\vc p',{\pmb\pi}')}
  \right)\,,
\end{equation}
and the next iteration starts with sampling a new momentum vector $(\vc p,{\pmb\pi})$.

@@

\section*{Acknowledgement}
This work was partly financed by the Eawag Discretionary Fund.
